Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-07-21 Writing/NatureWater/1stRoundRevision/replies/reply_to_reviewer_2.rmd

+ Though I am not trained as a hydrologist, my understanding is that aquifer saturation thickness is a good measure of general water availability in an aquifer but not necessarily how much water is available at a specific location or point. If this is the case, the authors should make clear why this is an appropriate measure to use in their study and potential limitations of using this approach.

\textcolor{blue}{Tim? this is realted to one of the reviewer 1's comments}

+ It is unclear to me why county and time fixed effects were not used in every regression. In particular, it does not look like they were included in the regression that modeled irrigated area as a function of saturated thickness and water deficit. Yet, accounting for time-invariant differences across counties is important, especially because this is the regression that showed the largest effect of saturated thickness on a measure of agricultural production (irrigated area). Is there not enough variation within a county through time to pick up the effect of saturated thickness on irrigated area? This should all be made clearer, and a model that better captures the causal relationship between saturated thickness and area irrigated should be used if possible. If it's not possible, the limitations of the current approach should be made very clear and potential issues with inference.

\textcolor{blue}{Thank you for this comment, which we believe led to a slighly better model and acknowledgin important limitations of our methods. County and year fixed effects were included in equation (1), but only year fixed effects were included in equation (2). As you gueesed, the problem of including county FEs lies in the massive elimination of variation in aquifer thickness. Please see Figure 6 in Appendix A for the histogram of (1) aquifer thickness demeaned by the county average, (2) aquifer thickness demeaned by the state average, and (3) aquifer thickness without demeaning. As you can see in the figure, inclusion of county FEs (mathematically equivalent to regression using county-demeaned aquifer thickness) dramatically reduces the total variation in aquifer thickness and not enough variations are left for accurately estimating its impact of the share of irrigated acres. The inclusion of State FEs does not diminish the variation in aquifer thickness. Therefore, while the inclusion of State FEs certainly does not provide as tight of a control as county FEs, we have decided to include State FEs for both corn and soybean irrigated share regrssions as it is better than not including them. Figure 7 compares the estiamted impacts of aquifer thickness with or without State FEs. The negative impact of reduction in aquifer thickness on the share of irrigated acres is slightly greater when state fixed effects are included. Therefore, it may be that aquifer thickness is not very highly correlated with important unobservable variables that are affecting irrigated share significantly. Given the direction of potential bias, providing tighter controls with county fixed effects may result in even more significant drop in the share of irrigated acres (we may be underestimating the impact of aquifer thickness.). All the discussion we are providing here is now seen in Appenddix A.
}

### 2023-07-21 Writing/NatureWater/1stRoundRevision/revised_manuscript/manuscript_DR (LAGAE-6N57C7R9D's conflicted copy 2023-07-19).tex
Response to reviewers added

Differentiating irrigated yields by the level of aquifer thickness alters these responses in two ways. First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \ref{fig:yield-response}). This reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers ability to fully meet crop water requirements during periods of more severe or extended precipitation deficits \citep{rad2020effects, foster2014modeling, hrozencik2017heterogeneous}. Second, for negative water deficits, counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \ref{fig:yield-response} experience marginal declines in yields resulting in productivity close to rainfed yields. This counter-intuitive result may be due to the increased possibility of over-irrigation in regions with high aquifer thickness and well capacities \citep{foster2019assessing,gibson2017case,gibson2019benchmarking}. However, it is important to highlight that impacts of aquifer thickness on yields for negative deficits is much less pronounced than that for extreme droughts (i.e. high deficits), in particular for corn.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(WB\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0-33.3\%)\), \([33.3-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) represents the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient. To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\phi_t\)) are also included and standard error estimation is clustered by county.

To estimate the additional impact of aquifer thickness on the share of total acres that are irrigated in a county, a fractional logit model under the generalize additive model framework is used in which the impact of aquifer thickness, average water deficit, and their interactions are estimated without needing to assume any functional forms. This allows for flexible representation of the non-linear relationships between drought risks, aquifer conditions, and agricultural land use, including potential thresholds in farmers irrigated area choices under water supply constraints that have been noted in prior work (REF). The model is given in Equation \ref{equation2} below:

>>>>

Differentiating irrigated yields by the level of aquifer thickness alters these responses in two ways. First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \ref{fig:yield-response}). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers ability to fully meet crop water requirements during periods of more severe or extended precipitation deficits \citep{rad2020effects, foster2014modeling, hrozencik2017heterogeneous}. Second, at negative water deficits, counties with the second highest and highest levels of aquifer thickness (i.e., 2nd and 3rd quantile in Figure \ref{fig:yield-response}) experience lower yields than counties with the lowest level of aquifer thickness (1st quantile), resulting in productivity close to rainfed yields. This counter-intuitive result may be due to the increased possibility of over-irrigation in regions with high aquifer thickness and well capacities \citep{foster2019assessing,gibson2017case,gibson2019benchmarking}. However, it is important to highlight that impacts of aquifer thickness on yields for negative deficits is much less pronounced than that for extreme droughts (i.e. high deficits), in particular for corn.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(WB\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0-33.3\%)\), \([33.3-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient. To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\phi_t\)) are also included and standard error estimation is clustered by county.

To estimate the additional impact of aquifer thickness on the share of total acres that are irrigated in a county, a fractional logit model under the generalized additive model framework is used in which the impact of aquifer thickness, average water deficit, and their interactions are estimated without needing to assume any functional forms. This allows for flexible representation of the non-linear relationships between drought risks, aquifer conditions, and agricultural land use, including potential thresholds in farmers irrigated area choices under water supply constraints that have been noted in prior work (REF). The model is given in Equation \ref{equation2} below:
